---
title: "CaseATAC: benchmarking within- and cross-omics annotation"
output: html_notebook
---

This case study assesses the hard classification performances of PhiSpace. This invovles

- Benchmarking within-modality annotation using scRNA-seq

- Benchmarking cross-modality annotation using 10x multiome


Note that `dat_dir` is the directory where you store the downloaded data and output from the R scripts. This directory should have the following structure:
```
dat_dir
│   
└───data
│   │
│   └───stemformatics
│   |   │   ref_dc.rds
│   |   │   ...
|   |   
│   └───Covid
|   |...
|   |
└───output
    │  
    └───CaseDC
    |
    └───CaseCITE
    |
    └───CaseATAC
    |...
```

```{r}
suppressPackageStartupMessages(library(PhiSpace))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(RColorBrewer))

dat_dir <- "/home/unimelb.edu.au/jmao1/PhiSpace/" # Replace this by your own directory 
source("utils.R")
```


See see `PrepareData.R` for data processing and normalisation.



# Within-omics 
Load benchmarking results using NeurIPS2021 multiome data (RNA part only). See `Case_Benchmark/` for R code for benchmarking. These objects can be downloaded from [OneDrive](https://unimelbcloud-my.sharepoint.com/:f:/g/personal/jiadong_mao_unimelb_edu_au/EqAiHIQq0plHsNs8gJWa9ucBFWKyxYGGtTE39UQ1HxQGIg?e=kM4SQE)
```{r}
# SingleR
SR_ann_res <- readRDS(
  paste0(dat_dir, "output/scRNAseq/NeurIPS2021_multiome_SR_ann_res_cor.rds")
)
# Seurat V3
Seu_ann_res <- readRDS(
  paste0(dat_dir, "output/scRNAseq/NeurIPS2021_multiome_Seu_ann_res_cor.rds")
)
# PhiSpace
# Original full Phi_ann_res object too large here we use a diet version
Phi_ann_res_path <- paste0(
  dat_dir,
  "output/scRNAseq/NeurIPS2021_multiome_Phi_ann_res_cor_diet.rds"
)
if(!file.exists(Phi_ann_res_path)){
  
  Phi_ann_res <- readRDS(
    paste0(
      dat_dir, "output/scRNAseq/NeurIPS2021_multiome_Phi_ann_res_cor.rds"
    )
  )
  
  Phi_ann_res_diet <- lapply(
    Phi_ann_res,
    function(Phi_ann_res_i){
      
      lapply(
        Phi_ann_res_i,
        function(x) x$errSR
      )
    }
  )
  
  saveRDS(
    Phi_ann_res_diet, Phi_ann_res_path
  )
} else {
  
  Phi_ann_res <- readRDS(Phi_ann_res_path)
}
scANVI_ann_res <- readRDS(
  paste0(dat_dir, "output/scRNAseq/NeurIPS2021_multiome_scANVI_ann_res_cor.rds")
)
```

Viz the errors: preparation.
```{r}
methodNumber <- 4
batchNames <- names(SR_ann_res)
errs <- matrix(
  NA, nrow = 0, ncol = 2
) 
temp <- vector("list", length(batchNames))
for(refIdx in 1:length(batchNames)){
  
  refBatchName <- batchNames[refIdx]
  queryBatchNames <- setdiff(batchNames, refBatchName)
  
  temp[[refIdx]] <- queryBatchNames
  
  for(queryIdx in 1:length(queryBatchNames)){
    
    queryBatchName <- queryBatchNames[queryIdx]
    errs <- rbind(
      errs, 
      SR_ann_res[[refBatchName]][[queryBatchName]]$errSR,
      Seu_ann_res[[refBatchName]][[queryBatchName]]$errSR,
      Phi_ann_res[[refBatchName]][[queryBatchName]],
      scANVI_ann_res[[refBatchName]][[queryBatchName]]$errSR
    )
  }
}
temp <- do.call(c, temp)
errs <- errs %>%
  as.data.frame() %>%
  `colnames<-`(
    c("overall", "balanced")
  ) %>%
  mutate(
    ref = rep(
      batchNames,
      rep((length(batchNames) - 1)*methodNumber, length(batchNames))
    ),
    query = rep(
      temp,
      rep(methodNumber, length(temp))
    ),
    method = rep(
      c("SingleR", "Seurat", "PhiSpace", "scANVI"),
      length(temp)
    )
  ) %>%
  pivot_longer(
    c(overall, balanced),
    names_to = "errorType",
    values_to = "error"
  )
```



Boxplot
```{r}
methodCols <- c(
  "#E41A1C", "#377EB8", "#984EA3", "#4DAF4A"
) %>%
  `names<-`(c("PhiSpace", "Seurat", "scANVI", "SingleR"))
p <- errs %>%
  filter(errorType == "balanced") %>%
  mutate(
    method = factor(
      method,
      levels = c("PhiSpace", "Seurat", "scANVI", "SingleR")
    )
  ) %>%
  ggplot(
    aes(
      x = method,
      y = error
    )
  ) +
  geom_boxplot(
    aes(
      fill = method
    ),
    linewidth = 0.3,
    outlier.size = 0.5
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Balanced error") +
  scale_fill_manual(
    values = methodCols,
    labels = c("PhiSpace", "SeuV3", "scANVI", "SingleR")
  ) +
  theme(
    axis.text.y = element_text(size = fsize),
    axis.title = element_text(size = fsize),
    legend.text = element_text(size = fsize, face = "bold"),
    legend.title = element_text(size = fsize, face = "bold"),
    legend.position = "right",
    legend.title.position = legend.title.position,
    legend.key = element_blank(),
    legend.spacing.x = unit(legend.spacing.x, "pt"),
    legend.key.spacing = unit(legend.key.spacing, "pt"),
    legend.key.spacing.y = unit(0, "pt"),
    legend.box.margin = margin(
      legend.box.margin,
      legend.box.margin,
      legend.box.margin,
      legend.box.margin
    ),
    legend.box.spacing = unit(legend.box.spacing, "pt"),
    legend.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      title = element_blank()
    )
  ) #+
  # facet_wrap(~errorType, nrow = 1, scales = "free") 
ggsave(
  "figs/classifErrs.pdf",
  p,
  width = 2.35,
  height = 2
)
```



# Cross-omics annotation


## Reference mapping
```{r}
bridgeATAC_list <- readRDS(paste0(dat_dir, "data/NeurIPS2021/bridgeATAC_list_cor.rds"))
  PhiSpaceAnn_cv <- readRDS(paste0(dat_dir, "output/CaseATAC/PhiSpaceAnn_cv_genAct_normPhiSc_cor.rds"))
  assay2use <- "logcounts"
reference <- readRDS(paste0(dat_dir, "data/NeurIPS2021/obj.rna_for_refMap_sce.rds"))
# Seurat BridgeInt annotations
bridgeIntAnn_cv <- readRDS(paste0(dat_dir, "output/CaseATAC/bridgeIntAnn_cv_cor.rds"))
originAnn_l <- readRDS(paste0(dat_dir, "output/CaseATAC/originAnn_l.rds"))
cellTypeTable <- readRDS(paste0(dat_dir, "output/CaseATAC/cellTypeTable.rds"))
```

Some numbers for paper.
```{r}
sapply(bridgeATAC_list, ncol) %>% sum
length(unique(reference$celltype.l2))
```


Bridge and query batches.
```{r}
# Bridge
(batchNames <- names(bridgeATAC_list))
bridgeBatchName <- "s4d8"
queryBatchNames <- c(
  "s1d1", "s1d2", "s2d1", "s2d4", "s3d3", "s3d6", "s4d1", "s4d9"
)
```

Rename cell types.
```{r}
refTypes_l2 <- unique(reference$celltype.l2) %>% sort
refTypes_l1 <- sapply(
  refTypes_l2,
  function(x){
    unique(reference$celltype.l1[reference$celltype.l2 == x])
  }
)
ref_lookup <- data.frame(
  l2 = refTypes_l2,
  l1 = refTypes_l1
)
## Query original annotations (ILC is hard to classify)
queryTypes_l2 <- rownames(cellTypeTable) %>% sort
queryTypes_l1 <- c(
  "B cell", "Mono/DC", "Mono/DC", "T cell", "T cell",
  "T cell", "T cell", "Mono/DC", "Progenitor cells", "Progenitor cells",
  "Progenitor cells", "Progenitor cells", "ILC", "Progenitor cells", "Progenitor cells",
  "B cell", "NK", "Progenitor cells", "Mono/DC", "B cell",
  "Progenitor cells", "B cell"
)
query_lookup <- data.frame(
  l2 = queryTypes_l2,
  l1 = queryTypes_l1
)
```


Subsampling 1000 cells from each batch.
```{r}
set.seed(39486723)
subN <- 1000
subIdx <- lapply(
  1:length(queryBatchNames),
  function(x){
    
    queryBatchName <- queryBatchNames[x]
    idx <- sample(
      1:ncol(bridgeATAC_list[[queryBatchName]]),
      subN
    )
    
    return(idx)
  }
)
comboATAC <- lapply(
  1:length(queryBatchNames),
  function(x){
    
    queryBatchName <- queryBatchNames[x]
    queryBatch <- bridgeATAC_list[[queryBatchName]]
    idx <- subIdx[[x]]
    queryBatch <- queryBatch[,idx]
    return(
      queryBatch
    )
  }
)

comboATAC <- Reduce(SingleCellExperiment::cbind, comboATAC)

comboATAC$cellType.main <- query_lookup$l1[match(comboATAC$cellType, query_lookup$l2)]

(ncomp <- length(unique(reference$celltype.l2)))
```

Compute UMAP using log normalised data.
```{r}
query_pc_umap <- computUMAP(
  t(assay(comboATAC, "logcounts")), 
  computPC = T, 
  ncomp = ncomp
)
```



PhiSpace UMAPs.
```{r}
YqueryHat_nrom <- do.call(
  rbind,
  lapply(
    1:length(queryBatchNames),
    function(x){
      
      queryBatchName <- queryBatchNames[x]
      PhiSc <- PhiSpaceAnn_cv[[bridgeBatchName]][[queryBatchName]]
      idx <- subIdx[[x]]
      
      return(PhiSc$queryScore_norm[idx,])
    }
  )
)
query_phi_umap <- computUMAP(
  YqueryHat_nrom, 
  computPC = F
)
```


SeuBI UMAPs
```{r}
BI_sc <- do.call(
  rbind,
  lapply(
    1:length(queryBatchNames),
    function(x){
      
      queryBatchName <- queryBatchNames[x]
      BISc <- bridgeIntAnn_cv[[bridgeBatchName]][[queryBatchName]]
      idx <- subIdx[[x]]
      
      return(t(BISc$scores.l2[,idx]))
    }
  )
)
query_seu_umap <- computUMAP(
  BI_sc, 
  computPC = F
)
```


Generate figures.
```{r fig.asp=1.5}
ggsave(
  "figs/UMAP_ctype.pdf",
  ggarrange(
    tempPlot(query_pc_umap, labs = comboATAC$cellType.main, colSc = "Set2"),
    tempPlot(query_seu_umap, labs = comboATAC$cellType.main, colSc = "Set2"),
    tempPlot(query_phi_umap, labs = comboATAC$cellType.main, colSc = "Set2"),
    nrow = 1,
    common.legend = T
  ),
  width = 7,
  height = 2.5
)
ggsave(
  "figs/UMAP_batch.pdf",
  ggarrange(
    tempPlot(query_pc_umap, labs = comboATAC$batch, labName = "batch"),
    tempPlot(query_seu_umap, labs = comboATAC$batch, labName = "batch"),
    tempPlot(query_phi_umap, labs = comboATAC$batch, labName = "batch"),
    nrow = 1,
    common.legend = T
  ),
  width = 7,
  height = 2.5
)
```




## Cell type prediction

Load PhiSpace cell typing results.
```{r}
bridgeIntAnn_cv <- readRDS(paste0(dat_dir, "output/CaseATAC/bridgeIntAnn_cv_cor.rds"))
bridgeIntAnnGeneAct_cv <- readRDS(paste0(dat_dir, "output/CaseATAC/bridgeIntAnnGeneAct_cv_cor.rds"))
PhiSpaceAnn_cv <- readRDS(
  paste0(
    dat_dir,
    "output/CaseATAC/PhiSpaceAnn_cv_genAct_normPhiSc_cor.rds"
  )
)
PhiSpaceAnn_cv_peaks <- readRDS(
  paste0(
    dat_dir,
    "output/CaseATAC/PhiSpaceAnn_cv_TF-IDF_cor.rds"
  )
)
```


```{r}
# List of original annotations
originAnn_l <- readRDS(paste0(dat_dir, "output/CaseATAC/originAnn_l.rds"))
cellTypeTable <- readRDS(paste0(dat_dir, "output/CaseATAC/cellTypeTable.rds"))
# YrefHats <- readRDS(paste0(dat_dir, "output/CaseATAC/YrefHats.rds"))
reference <- readRDS(paste0(dat_dir, "data/NeurIPS2021/obj.rna_for_refMap_sce.rds"))
refTypes_l2 <- unique(reference$celltype.l2) %>% sort
refTypes_l1 <- sapply(
  refTypes_l2,
  function(x){
    unique(reference$celltype.l1[reference$celltype.l2 == x])
  }
)
ref_lookup <- data.frame(
  l2 = refTypes_l2,
  l1 = refTypes_l1
)
## Query original annotations (ILC is hard to classify)
queryTypes_l2 <- rownames(cellTypeTable) %>% sort
queryTypes_l1 <- c(
  "B cell", "Mono/DC", "Mono/DC", "T cell", "T cell",
  "T cell", "T cell", "Mono/DC", "Progenitor cells", "Progenitor cells",
  "Progenitor cells", "Progenitor cells", "ILC", "Progenitor cells", "Progenitor cells",
  "B cell", "NK", "Progenitor cells", "Mono/DC", "B cell",
  "Progenitor cells", "B cell"
)
query_origin_lookup <- data.frame(
  l2 = queryTypes_l2,
  l1 = queryTypes_l1
)
```


```{r}
nBatches <- length(bridgeIntAnn_cv)
bridgeNames <- names(bridgeIntAnn_cv)
errs_l <- vector("list", nBatches*(nBatches - 1))
for(jj in 1:length(bridgeNames)){
  
  bridgeName <- bridgeNames[jj]
  bridgeIntAnn_l <- bridgeIntAnn_cv[[bridgeName]]
  bridgeIntAnn_l_geneAct <- bridgeIntAnnGeneAct_cv[[bridgeName]]
  PhiSpaceAnn_l <- PhiSpaceAnn_cv[[bridgeName]]
  PhiSpaceAnn_l_peaks <- PhiSpaceAnn_cv_peaks[[bridgeName]]
  queryNames <- names(bridgeIntAnn_l)
  
  for(kk in 1:length(queryNames)){
    
    queryName <- queryNames[kk]
    
    # Translate Seurat bridgeInt labels
    bridgeIntAnn <- bridgeIntAnn_l[[queryName]]$metadata$predicted.l2
    bridgeIntAnn_l1 <- ref_lookup$l1[match(bridgeIntAnn, ref_lookup$l2)]
    bridgeIntAnn_geneAct <- bridgeIntAnn_l_geneAct[[queryName]]$metadata$predicted.l2
    bridgeIntAnn_geneAct_l1 <- ref_lookup$l1[match(bridgeIntAnn_geneAct, ref_lookup$l2)]
    
    nNAs <- bridgeIntAnn_l1 %>% is.na() %>% sum()
    if(nNAs) stop("NAs appeared in converting l2 to l1 cell type lables.")
    
    # Translate PhiSpace labels
    PhiSpaceAnn <- PhiSpaceAnn_l[[queryName]]$queryScore_norm %>% getClass
    PhiSpaceAnn_l1 <- ref_lookup$l1[match(PhiSpaceAnn, ref_lookup$l2)]
    PhiSpaceAnnPeaks <- PhiSpaceAnn_l_peaks[[queryName]]$queryScore_norm %>% getClass()
    PhiSpaceANN_l1_peaks <- ref_lookup$l1[match(PhiSpaceAnnPeaks, ref_lookup$l2)]
    
    nNAs <- PhiSpaceAnn_l1 %>% is.na() %>% sum()
    if(nNAs) stop("NAs appeared in converting l2 to l1 cell type lables.")
    
    # Original annotations
    originAnn <- originAnn_l[[queryName]]$cellType
    originAnn_l1 <- query_origin_lookup$l1[
      match(originAnn, query_origin_lookup$l2)
    ]
    
    # Some cell types do not have good correspondence to l1 cell types
    types2move <- c("ILC")
    idx <- !(originAnn %in% types2move)
    
    
    ## PhiSpace errs by other methods
    
    
    errs_ind <- (jj-1)*nBatches + kk
    errs_l[[errs_ind]] <-
      data.frame(
        errs = c(
          PhiSpace:::classErr(bridgeIntAnn_l1[idx], originAnn_l1[idx])$err,
          PhiSpace:::classErr(bridgeIntAnn_geneAct_l1[idx], originAnn_l1[idx])$err,
          PhiSpace:::classErr(PhiSpaceAnn_l1[idx], originAnn_l1[idx])$err,
          PhiSpace:::classErr(PhiSpaceANN_l1_peaks[idx], originAnn_l1[idx])$err
        ),
        typeOfErr = c(
          "overall", "balanced", "overall", "balanced", 
          "overall", "balanced",  "overall", "balanced"
        ),
        bridge = rep(bridgeName, 8),
        query = rep(queryName, 8),
        method = c(
          "SeuratBridge", "SeuratBridge", 
          "SeuratBridgeGeneAct", "SeuratBridgeGeneAct",
          "PhiSpace", "PhiSpace",
          "PhiSpacePeaks", "PhiSpacePeaks"
        )
      )  
    
  }
}
```

Visualising the errors
```{r}
errs_long <- do.call(rbind, errs_l)
# BridgeInt vs PhiSpace (just shown balanced error here)
p <- errs_long %>%
  filter(typeOfErr == "balanced") %>%
  mutate(
    method = factor(
      method,
      levels = c("PhiSpace", "PhiSpacePeaks",  "SeuratBridgeGeneAct", "SeuratBridge")
    ) %>%
    `levels<-`(
      c("Phi_GA", "Phi_Peaks", "SeuBI_GA", "SeuBI_Peaks")
    )
  )  %>%
  ggplot(aes(method, errs, fill = method)) +
  theme_bw() +
  geom_boxplot(
    linewidth = 0.3,
    outlier.size = 0.5
  ) +
  scale_fill_manual(
    values = c(
      "#E41A1C", "#FB8072", "#377EB8", "#80B1D3"
    ) %>%
      `names<-`(
        c("Phi_GA", "Phi_Peaks", "SeuBI_GA", "SeuBI_Peaks")
      )
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  ylab("Balanced error") +
  theme(
    axis.text = element_text(size = fsize),
    axis.title = element_text(size = fsize),
    legend.text = element_text(size = fsize, face = "bold"),
    legend.title = element_text(size = fsize, face = "bold"),
    legend.position = "right",
    legend.title.position = legend.title.position,
    legend.key = element_blank(),
    legend.spacing.x = unit(legend.spacing.x, "pt"),
    legend.key.spacing = unit(legend.key.spacing, "pt"),
    legend.key.spacing.y = unit(0, "pt"),
    legend.box.margin = margin(
      legend.box.margin,
      legend.box.margin,
      legend.box.margin,
      legend.box.margin
    ),
    legend.box.spacing = unit(legend.box.spacing, "pt"),
    legend.background = element_blank()
  ) +
  guides(
    fill = guide_legend(
      title = element_blank()
    )
    # override.aes = list(
    #     size = legend.obj.size
    # )
  ) #+
  # facet_wrap(~typeOfErr, nrow = 1, scales = "free") 
  
ggsave(
  "figs/classErr.pdf",
  p,
  width = 2.5,
  height = 2
)
```

## Semi-supervised clustering
Organise results.
```{r}
clustMeas_path <- paste0(
  dat_dir,
  "output/CaseATAC/clustMeas.rds"
)
if(!file.exists(clustMeas_path)){
  
  bridgeBatchNames <- readRDS(paste0(dat_dir, "output/CaseATAC/cellTypeTable.rds")) %>%
    colnames()
  clust_res <- readRDS(paste0(dat_dir, "output/CaseATAC/clus_res_cor.rds"))
  clust_res_alt <- readRDS(paste0(dat_dir, "output/CaseATAC/clus_res_cor_PCAonly_stack.rds"))
  clust_res_peaks <- readRDS(paste0(dat_dir, "output/CaseATAC/clus_res_cor_PeaksOnly_stack.rds"))
  names(clust_res) <- names(clust_res_alt) <- names(clust_res_peaks) <- bridgeBatchNames
  clustMeas <- vector('list', length(bridgeBatchNames)) %>%
    `names<-`(bridgeBatchNames)
  for(ii in 1:length(bridgeBatchNames)){
    
    bridgeBatchName <- bridgeBatchNames[ii]
    clustRes <- clust_res[[bridgeBatchName]]
    
    origin <- clustRes[[1]]
    origin_l1 <- clustRes[[2]]
    originClust <- clustRes[[3]]
    originClust_l1 <- clustRes[[4]]
    clustRes <- clustRes[[5]]
    
    clustRes[[1]] <- c(
      clustRes[[1]],
      clust_res_alt[[bridgeBatchName]][[5]][[1]],
      clust_res_peaks[[bridgeBatchName]][[5]][[1]]
    )
    clustRes[[2]] <- c(
      clustRes[[2]],
      clust_res_alt[[bridgeBatchName]][[5]][[2]],
      clust_res_peaks[[bridgeBatchName]][[5]][[2]]
    )
    
    meas <- lapply(
      1:length(clustRes[[1]]),
      function(x){
        # All measures implemented in aricode package
        clustcomp_l1 <- suppressWarnings(
          aricode::clustComp(clustRes[[1]][[x]], originClust_l1)
        )
        clustcomp_l2 <- suppressWarnings(
          aricode::clustComp(clustRes[[2]][[x]], originClust)
        )
        # Purity
        Purs_l1 <- purity(clustRes[[1]][[x]], originClust_l1)
        Purs_l2 <- purity(clustRes[[2]][[x]], originClust)
        Pur_l1 <- Purs_l1$Pur
        Pur_l2 <- Purs_l2$Pur
        InvPur_l1 <- Purs_l1$invPur
        InvPur_l2 <- Purs_l2$invPur
        Fmeasure_l1 <- Purs_l1$Fmeasure
        Fmeasure_l2 <- Purs_l2$Fmeasure
        
        l1Purs <- list(
          Pur = Pur_l1,
          InvPur = InvPur_l1,
          Fmeasure = Fmeasure_l1
        )
        l2Purs <- list(
          Pur = Pur_l2,
          InvPur = InvPur_l2,
          Fmeasure = Fmeasure_l2
        )
        
        return(
          list(
            l1 = c(
              l1Purs,
              clustcomp_l1
            ),
            l2 = c(
              l2Purs,
              clustcomp_l2
            )
          )
        )
        
        
      }
    )
    
    names(meas) <- c(
      "TFIDF_renorm", 
      "PC_renorm", 
      "Phi", 
      "PhiPeaks", 
      "BI", 
      "BIgenAct", 
      "PC", 
      "TFIDF"
    )
    
    clustMeas[[ii]] <- meas
  }
  
  saveRDS(clustMeas, clustMeas_path)
} else {
  
  clustMeas <- readRDS(clustMeas_path)
}
```



Visualisation.
```{r}
methNames <- c(
  # "TFIDF_renorm",
  "TFIDF", 
  # "PC_renorm", 
  "PC", 
  "Phi", 
  "PhiPeaks",
  "BI", 
  "BIgenAct"
)
out <- vector("list", length(methNames))
for(ii in 1:length(methNames)){
  
  methName <- methNames[ii]
  
  out_l1 <- sapply(
   clustMeas,
   function(x){
     unlist(x[[methName]]$l1)
   }
  )
  out_l2 <- sapply(
    clustMeas,
    function(x){
      unlist(x[[methName]]$l2)
    }
  )
  
  out_l1 <- out_l1 %>%
    as.data.frame() %>%
    mutate(measName = rownames(.)) %>%
    pivot_longer(!measName, names_to = "batch", values_to = "value") %>%
    mutate(
      resolution = rep("l1", nrow(.)),
      clustMethod = rep(methName, nrow(.))
    )
  out_l2 <- out_l2 %>%
    as.data.frame() %>%
    mutate(measName = rownames(.)) %>%
    pivot_longer(!measName, names_to = "batch", values_to = "value") %>%
    mutate(
      resolution = rep("l2", nrow(.)),
      clustMethod = rep(methName, nrow(.))
    )
  
  out[[ii]] <- rbind(
    out_l1,
    out_l2
  )
}

plot_dat <- do.call(rbind, out)
methodCols <- c(
  "#E41A1C", "#377EB8", 
  "#FB8072", "#80B1D3", 
  "#4DAF4A", "#984EA3", 
  "#4DAF4A", "#984EA3"
) %>%
  `names<-`(
    c(
      "Phi-GA", "SeuBI-GA", 
      "Phi-Peaks", "SeuBI-Peaks", 
      "GA-renorm", "Peaks-renorm", 
      "GA", "Peaks"
    )
  )
tempPlot <- function(ClustTruth = "l1", omit.strip.text = FALSE){
  
  showMeas <- c(
    # "RI", 
    "ARI", 
    # "MI", 
    "NMI", 
    # "Pur",
    # "InvPur"
    "Fmeasure"
  )
  
  ytitle <- ifelse(
    ClustTruth == "l1",
    "Nclust = 6",
    "Nclust = 22"
  )
  
  p <- plot_dat %>%
    filter((resolution == ClustTruth) & (measName %in% showMeas) ) %>%
    mutate(
      clustMethod = factor(
          clustMethod,
          levels = c(
            "Phi", "PhiPeaks", "BIgenAct", "BI", "PC", "TFIDF", "PC_renorm", "TFIDF_renorm"
          )
        ) %>%
        `levels<-`(
          c(
            "Phi-GA", "Phi-Peaks", 
            "SeuBI-GA", "SeuBI-Peaks", 
            "GA", "Peaks",
            "GA-renorm", "Peaks-renorm"
          )
        ),
      measName = factor(
        measName,
        levels = showMeas,
        labels = c(
          # "Rand index", 
          "Adjusted RI",
          # "Mutual info", 
          "Normalised MI",
           "F measure"
          # "InvPurity"
        )
      )
    ) %>%
    ggplot(
      aes(
        clustMethod, 
        value, 
        fill = clustMethod
      )
    ) +
    geom_boxplot(
      linewidth = 0.3,
      outlier.size = 0.5
    ) +
    theme_bw() +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    ) +
    labs(y = ytitle) +
    scale_fill_manual(
      values = methodCols,
      name = "ClustMethod"
    ) +
    theme(
      axis.text.y = element_text(size = fsize),
      axis.title = element_text(size = fsize),
      legend.text = element_text(size = fsize, face = "bold"),
      legend.title = element_text(size = fsize, face = "bold"),
      legend.position = legend.position,
      legend.title.position = legend.title.position,
      legend.key = element_blank(),
      legend.spacing.x = unit(legend.spacing.x, "pt"),
      legend.key.spacing = unit(legend.key.spacing, "pt"),
      legend.key.spacing.y = unit(legend.key.spacing.y, "pt"),
      legend.box.margin = margin(
        legend.box.margin,
        legend.box.margin,
        legend.box.margin,
        legend.box.margin
      ),
      legend.box.spacing = unit(legend.box.spacing, "pt"),
      legend.background = element_blank()
    ) +
    guides(
      fill = guide_legend(
        title = element_blank(),
        nrow = 1
      )
      # override.aes = list(
      #     size = legend.obj.size
      # )
    ) +
    facet_wrap(~measName, nrow = 1, scales = "free") 
  
  if(omit.strip.text){
    p <- p + 
      theme(
        strip.text = element_blank()
      )
  } else {
    p <- p +
      theme(
        strip.text = element_text(size = fsize)
      )
  }
}
ggsave(
  filename = "figs/ClustRes_box.pdf", 
  plot = ggarrange(tempPlot(), tempPlot("l2",T), nrow = 2, common.legend = T), 
  width = 4.7,
  height = 4
)
```
