---
title: "CaseCITE: projecting Vento-Tormo CITE-seq to Haniffa CITE-seq"
output: html_notebook
---

**Bibliographic note**: Vento-Tormo Lab published their analysis of their autoimmunity Covid dataset [here](https://onlinelibrary.wiley.com/doi/10.1002/eji.202350633). Biological background knowledge referred to here is mostly from that reference.

Note that `dat_dir` is the directory where you store the downloaded data and output from the R scripts. This directory should have the following structure:
```
dat_dir
│   
└───data
│   │
│   └───stemformatics
│   |   │   ref_dc.rds
│   |   │   ...
|   |   
│   └───Covid
|   |...
|   |
└───output
    │  
    └───CaseDC
    |
    └───CaseCITE
    |
    └───CaseATAC
```

```{r}
suppressPackageStartupMessages(library(PhiSpace))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(RColorBrewer))

dat_dir <- "/home/unimelb.edu.au/jmao1/PhiSpace/" # Replace this by your own directory 
source("utils.R")
```

# Get PhiSpace annotation results

Data can be downloaded from [OneDrive](https://unimelbcloud-my.sharepoint.com/:f:/g/personal/jiadong_mao_unimelb_edu_au/EimPelXY7CpHqNKd4vKfLQABhDNcmkpQtzlm2uKOQMGtPQ?e=uQhpBR)

See [CaseCITE_prepareData.Rmd]{.underline} for data preparation.

## ADT modality

Several issues specific to ADT modality:

-   Naming convention of ADTs different across datasets: see [CaseCITE_prepareData.Rmd]{.underline} for details.

-   ADT assay has limited number of features, which may not be enough to learn too many cell type labels.

```{r}
PhiResPath <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoADTPhiRes.rds"
  )
reference <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Haniffa/HaniffaADT.rds"
    )
  )
# See Case_CITE_doubleCovidCellType.Rmd for renaming of ADTs
query <- readRDS(
  paste0(
    dat_dir,
    "data/Covid/Vento-Tormo/ADT.rds"
  )
)

if(!file.exists(PhiResPath)){
  PhiSpaceAssay <- "data"
  YtrainName <- c("initial_clustering", "Status_on_day_collection_summary")
  PhiRes <- PhiSpaceR_1ref(
    reference, 
    query = query, 
    phenotypes = YtrainName, 
    PhiSpaceAssay = PhiSpaceAssay,
    regMethod = "PLS", 
    scale = FALSE
  )
  saveRDS(
    PhiRes, PhiResPath
  )
  gc()
} else {
  
  PhiResADT <- readRDS(PhiResPath)
}
```

Define some parameters that will be useful.

```{r}
severityNames <- unique(reference$Status_on_day_collection_summary %>% as.character())
autoimmuneNames <- unique(query$Group %>% as.character())
```


## RNA modality

Run PhiSpace for RNA asssay
```{r}
PhiResPath <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoRNAPhiRes.rds"
)
if(!file.exists(PhiResPath)){
  
  reference <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Haniffa/HaniffaRNA.rds"
    )
  )
  query <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Vento-Tormo/RNA.rds"
    )
  )
  
  PhiSpaceAssay <- "logcounts"
  YtrainName <- c("full_clustering", "Status_on_day_collection_summary")
  PhiRes <- PhiSpaceR_1ref(
    reference, 
    query = query, 
    phenotypes = YtrainName, 
    PhiSpaceAssay = PhiSpaceAssay,
    regMethod = "PLS"
  )
  saveRDS(
    PhiRes, PhiResPath
  )
  gc()
} else {
  
  PhiResRNA <- readRDS(PhiResPath)
}
``` 


# Batch effects in query

Load original UMAP for query data (batch-corrected).

```{r}
ventoRNAumap <- readRDS(
  paste0(dat_dir, "data/Covid/Vento-Tormo/ventoRNAumap.rds")
)
```


Compute UMAP using PhiSpace embeddings.

```{r}
set.seed(9183247)
idx <- sample(1:ncol(query), 10000)
quSubsetIdx <- idx # Store this object for future reference

pathADT <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoADTPhiUMAPquery.rds"
  )
pathRNA <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoRNAPhiUMAPquery.rds"
)
pathCombo <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoComboPhiUMAPquery.rds"
)

YrefHatRNA_norm <- normPhiScores(PhiResRNA$YrefHat)
YrefhatADT_norm <- normPhiScores(PhiResADT$YrefHat)
PhiScRNA_norm <- normPhiScores(PhiResRNA$PhiSpaceScore)
PhiScADT_norm <- normPhiScores(PhiResADT$PhiSpaceScore)

if(!(file.exists(pathADT) & file.exists(pathRNA) )){
  
  umapPhiRNA_res <- umap::umap(
    PhiScRNA_norm[idx,]
  )
  umapPhiADT_res <- umap::umap(
    PhiScADT_norm[idx,]
  )
  
  saveRDS(
    umapPhiADT_res, pathADT
  )
  saveRDS(
    umapPhiRNA_res, pathRNA
  )
} else {
  
  umapPhiADT_res <- readRDS(pathADT)
  umapPhiRNA_res <- readRDS(pathRNA)
}



# Concatenate two annotations
if(!file.exists(pathCombo)){
  
  comboPCA_res <- getPC(
    cbind(PhiScADT_norm, PhiScRNA_norm)[idx,],
    ncomp = 64 # number of latent variables used in 
  )
  
  umapPhiCombo_res <- umap::umap(
    comboPCA_res$scores
  )
  
  saveRDS(
    umapPhiCombo_res,
    pathCombo
  )
} else {
  
  umapPhiCombo_res <- readRDS(
    pathCombo
  )
}
```

Compute UMAP using X-space PCs.

```{r}
pathADT <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoADTomicsUMAPquery.rds"
  )
pathRNA <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoRNAomicsUMAPquery.rds"
)

if(!(file.exists(pathADT) & file.exists(pathRNA) )){
  
  quRNA <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Vento-Tormo/RNA.rds"
    )
  )
  quADT <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Vento-Tormo/ADT.rds"
    )
  )

  
  pca_adt <- getPC(
    t(assay(quADT, "data")[,quSubsetIdx]),
    ncomp = ncol(PhiScADT_norm)
  )
  pca_rna <- getPC(
    t(assay(quRNA, "logcounts")[,quSubsetIdx]),
    ncomp = ncol(PhiScRNA_norm)
  )
  
  umapOmicsADT_res <- umap::umap(
    pca_adt$scores
  )
  umapOmicsRNA_res <- umap::umap(
    pca_rna$scores
  )
  
  saveRDS(
    umapOmicsADT_res, pathADT
  )
  saveRDS(
    umapOmicsRNA_res, pathRNA
  )
} else {
  
  umapOmicsADT_res <- readRDS(pathADT)
  umapOmicsRNA_res <- readRDS(pathRNA)
}
```


Compute UMAP using X-space reference mapping.

```{r}
pathADT <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoADTomicsMapUMAPquery.rds"
  )
pathRNA <- paste0(
    dat_dir,
    "output/CaseCITE/Covid/doublePhenoRNAomicsMapUMAPquery.rds"
)

if(!(file.exists(pathADT) & file.exists(pathRNA) )){
  
  refADT <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Haniffa/HaniffaADT.rds"
    )
  )
  refRNA <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Haniffa/HaniffaRNA.rds"
    )
  )
  quRNA <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Vento-Tormo/RNA.rds"
    )
  )
  quADT <- readRDS(
    paste0(
      dat_dir,
      "data/Covid/Vento-Tormo/ADT.rds"
    )
  )

  commonGenes <- intersect(
    rownames(refRNA),
    rownames(quRNA)
  )
  
  pca_adt <- getPC(
    t(assay(refADT, "data")),
    ncomp = ncol(PhiScADT_norm)
  )
  pca_rna <- getPC(
    t(assay(refRNA, "logcounts")[commonGenes,]),
    ncomp = ncol(PhiScRNA_norm)
  )
  
  mappedADT <- scale(
    t(assay(quADT, "data")[,quSubsetIdx]),
    center = T,
    scale = F
  ) %*%
    pca_adt$loadings 
  
  mappedRNA <- scale(
    t(assay(quRNA, "logcounts")[commonGenes,quSubsetIdx]),
    center = T,
    scale = F
  ) %*%
    pca_rna$loadings 
  
  
  umapOmicsADT_res <- umap::umap(
    mappedADT
  )
  umapOmicsRNA_res <- umap::umap(
    mappedRNA
  )
  
  saveRDS(
    umapOmicsADT_res, pathADT
  )
  saveRDS(
    umapOmicsRNA_res, pathRNA
  )
} else {
  
  umapOmicsMapADT_res <- readRDS(pathADT)
  umapOmicsMapRNA_res <- readRDS(pathRNA)
}
```



UMAPs using Seurat v3 ref mapping results. See `runSeurat.R` for details.
```{r}
pathCombo <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoComboSeuUMAPquery.rds"
)

SeuResPathRNA <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoRNASeuRes.rds"
)
SeuResPathADT <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoADTSeuRes.rds"
)
SeuResPathADTRef <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoADTSeuResRef.rds"
)
SeuResPathRNARef <- paste0(
  dat_dir,
  "output/CaseCITE/Covid/doublePhenoRNASeuResRef.rds"
)

SeuRNA <- readRDS(SeuResPathRNA)
SeuADT <- readRDS(SeuResPathADT)
SeuRNAref <- readRDS(SeuResPathRNARef)
SeuADTref <- readRDS(SeuResPathADTRef)

SeuRNA <- cbind(
  SeuRNA[[1]][,-c(1,ncol(SeuRNA[[1]]))],
  SeuRNA[[2]][,-c(1,ncol(SeuRNA[[2]]))]
)
SeuADT <- cbind(
  SeuADT[[1]][,-c(1,ncol(SeuADT[[1]]))],
  SeuADT[[2]][,-c(1,ncol(SeuADT[[2]]))]
)
SeuRNAref <- cbind(
  SeuRNAref[[1]][,-c(1,ncol(SeuRNAref[[1]]))],
  SeuRNAref[[2]][,-c(1,ncol(SeuRNAref[[2]]))]
)
SeuADTref <- cbind(
  SeuADTref[[1]][,-c(1,ncol(SeuADTref[[1]]))],
  SeuADTref[[2]][,-c(1,ncol(SeuADTref[[2]]))]
)


if(!file.exists(pathCombo)){
  
  SeuRNA_sub <- SeuRNA[quSubsetIdx, ]
  SeuADT_sub <- SeuADT[quSubsetIdx, ]

  comboPCA_res <- getPC(
    cbind(SeuRNA_sub, SeuADT_sub) %>% as.matrix(),
    ncomp = 64 # number of latent variables used in 
  )
  
  umapSeuCombo_res <- umap::umap(
    comboPCA_res$scores
  )
  
  saveRDS(umapSeuCombo_res, pathCombo)
} else {
  
  umapSeuCombo_res <- readRDS(pathCombo)
}
```






Plot the UMAPs.

```{r fig.asp=0.8}
PhiPredCellType = getClass(PhiScRNA_norm[,1:51])
PhiPredCellType[PhiPredCellType %in% c(
  "ASDC", "DC1", "DC2", "DC3", "pDC", "DC_prolif"
)] <- "DC"
PhiPredCellType[PhiPredCellType %in% c(
  "B_exhausted",           "B_immature",          
  "B_malignant",           "B_naive",               "B_non-switched_memory",
  "B_switched_memory"
)] <- "B cell"
PhiPredCellType[PhiPredCellType %in% c(
  "Plasma_cell_IgA", "Plasma_cell_IgG", "Plasma_cell_IgM","Plasmablast"
)] <- "B plasma"
PhiPredCellType[PhiPredCellType %in% c(
  "C1_CD16_mono",          "CD14_mono",            "CD16_mono",
  "CD83_CD14_mono", "Mono_prolif"
)] <- "Mono"
PhiPredCellType[PhiPredCellType %in% c(
  "CD4.CM",                "CD4.EM",               
  "CD4.IL22",              "CD4.Naive",             "CD4.Prolif",           
  "CD4.Tfh",               "CD4.Th1",               "CD4.Th17",             
  "CD4.Th2"
)] <- "CD4 T"
PhiPredCellType[PhiPredCellType %in% c(
  "CD8.EM",                "CD8.Naive",  "CD8.Prolif",            "CD8.TE"
)] <- "CD8 T"
PhiPredCellType[PhiPredCellType %in% c(
  "gdT", "MAIT", "Treg" 
)] <- "other T"
PhiPredCellType[PhiPredCellType %in% c(
  "NK_16hi",               "NK_56hi",               "NK_prolif",  "NKT"
)] <- "NK"
PhiPredCellType[PhiPredCellType %in% c(
  "HSC_CD38neg", "HSC_CD38pos",           "HSC_erythroid",         "HSC_MK",               
  "HSC_myeloid",           "HSC_prolif"
)] <- "HSC"
PhiPredCellType[PhiPredCellType %in% c(
  "ILC1_3", "ILC2", "Platelets", "RBC", "HSC"
)] <- "other"

# Main cell types for query
majCellType <- query$Annotation[idx] %>% as.character()
fineTypes <- sort(unique(as.character(query$Annotation)))
majTypes <- c(
  "Mono", "Mono", "CD4", "CD4",
  "CD8", "CD8", "DC", "DC",
  "DC", "B", "other", "B", 
  "other T", "B", "B", "other",
  "NK", "NK", "NK", "DC",
  "B", "other", "other", "NK",
  "other T", "other", "other T", "other T"
)
for(ii in 1:length(fineTypes)){
  ctype <- fineTypes[ii]
  majCellType[majCellType == ctype] <- majTypes[ii] 
}
table(majCellType)


# Simplified disease names
disease <- query$Group[idx] %>% as.character()
disNames <- sort(
  unique(query$Group %>% as.character())
)
simpDisease <- c(
  "Control", "MS", "Ps", "RA"
)
for(ii in 1:length(disNames)){
  dName <- disNames[ii]
  disease[disease == dName] <- simpDisease[ii] 
}

cols_disease <- c(
        Control = "gray",
        MS = "#377EB8",
        RA = "#4DAF4A",
        Ps = "#984EA3"
)
# cols_disease <- aurora_cols_dis
# names(cols_disease) <- c(
#   "Control", "MS", "RA", "Ps"
# )

cols_type <- brewer.pal(8, "Set1")
```



Calculate alignment score.

Using all cells and no PC (took long to run):
- Batch: 0.7823567 0.8242383
- Disease: 0.6227782 0.6815256
- Cell type: 0.4062466 0.3734180

Using 10,000 cells and 64 PCs (same as UMAP):


```{r}
sc_list <- list(
  cbind(PhiScADT_norm, PhiScRNA_norm)[idx,],
  cbind(SeuADT, SeuRNA)[idx,]
)

# Major cell type
sapply(
  sc_list,
  function(x){
    
    alignmentScore(
      x,
      batch = majCellType
    )
  }
)
lab <- "Batch"
sapply(
  sc_list,
  function(x){
    
    alignmentScore(
      x,
      batch = colData(query)[idx,lab]
    )
  }
)
lab <- "Group"
sapply(
  sc_list,
  function(x){
    
    alignmentScore(
      x,
      batch = colData(query)[idx,lab]
    )
  }
)
lab <- "Annotation"
sapply(
  sc_list,
  function(x){
    
    alignmentScore(
      x,
      batch = colData(query)[idx,lab]
    )
  }
)
```

Plot

```{r}
tempUMAPobj <- list(layout = ventoRNAumap[quSubsetIdx, ])
# Save plots
# RNA disease
colVar <- "disease"
p11 <- tempPlotUMAP(umapOmicsRNA_res, idx, colVar) %>% adjPlots()
p21 <- tempPlotUMAP(umapOmicsADT_res, idx, colVar) %>% adjPlots()
p31 <- tempPlotUMAP(tempUMAPobj, idx, colVar) %>% adjPlots()
p41 <- tempPlotUMAP(umapPhiCombo_res, idx, colVar) %>% adjPlots()
p51 <- tempPlotUMAP(umapSeuCombo_res, idx, colVar) %>% adjPlots()

# RNA celltype
colVar <- "celltype"
p12 <- tempPlotUMAP(umapOmicsRNA_res, idx, colVar) %>% adjPlots()
p22 <- tempPlotUMAP(umapOmicsADT_res, idx, colVar) %>% adjPlots()
p32 <- tempPlotUMAP(tempUMAPobj, idx, colVar) %>% adjPlots()
p42 <- tempPlotUMAP(umapPhiCombo_res, idx, colVar) %>% adjPlots()
p52 <- tempPlotUMAP(umapSeuCombo_res, idx, colVar) %>% adjPlots()

# RNA celltype
colVar <- "batch"
p13 <- tempPlotUMAP(umapOmicsRNA_res, idx, colVar) %>% adjPlots()
p23 <- tempPlotUMAP(umapOmicsADT_res, idx, colVar) %>% adjPlots()
p33 <- tempPlotUMAP(tempUMAPobj, idx, colVar) %>% adjPlots()
p43 <- tempPlotUMAP(umapPhiCombo_res, idx, colVar) %>% adjPlots()
p53 <- tempPlotUMAP(umapSeuCombo_res, idx, colVar) %>% adjPlots()

ggsave(
  "figs/umap1.pdf", 
  ggarrange(p11, p12, p13, nrow = 1, common.legend = F), 
  width = 5, height = 1.7
)
ggsave(
  "figs/umap2.pdf", 
  ggarrange(p21, p22, p23, nrow = 1, legend = F), 
  width = 5, height = 1.5
)
ggsave(
  "figs/umap3.pdf", 
  ggarrange(p31, p32, p33, nrow = 1, legend = F), 
  width = 5, height = 1.5
)
ggsave(
  "figs/umap4.pdf", 
  ggarrange(p41, p42, p43, nrow = 1, legend = F), 
  width = 5, height = 1.5
)
ggsave(
  "figs/umap5.pdf", 
  ggarrange(p51, p52, p53, nrow = 1, legend = F), 
  width = 5, height = 1.5
)
```






# Interpretation of continuous phenotyping

Existing methods typically annotate the cells and calculate cell type proportions for each disease condition. This method is highly sensitive to how the annotation is done. As a result, the direct quantitative comparison between cell type compositions derived in two studies may not always be possible.

We solve this problem by calculating the total/median/mean cell type scores by disease conditions, in both reference and query.

```{r fig.asp=1}
library(ComplexHeatmap)
library(seriation)
typeNames <- sort(
  colnames(YrefHatRNA_norm)[1:51]
)
ADTtypeNames <- sort(
  colnames(YrefhatADT_norm)[1:18]
)



refLabs <- reference$Status_on_day_collection_summary
sev_lvls <- levels(refLabs)
sev_lvls_simp <- c(
  "Asymp", "Critical", "Healthy", "LPS10h", "LPS90m",
  "Mild", "Moderate", "NonCovid", "Severe"  
)
levels(refLabs) <- sev_lvls_simp
condNames <- c(
  "Healthy", "LPS90m", "LPS10h", "NonCovid", "Asymp",
  "Mild", "Moderate", "Severe", "Critical"
)
refLabs <- factor(
  refLabs,
  levels = condNames
)
quLabs <- query$Group
dis_lvls <- levels(quLabs)
dis_lvls_simp <- c(
  "Control", "MS", "Ps", "RA"
)
levels(quLabs) <- dis_lvls_simp

disNames <- c(
  "Healthy", "Asymp", "Mild", "Moderate", "Severe", "Critical"
)

sc_RNA_ref <- mk_sc(YrefHatRNA_norm, typeNames, refLabs)[,disNames]
sc_ADT_ref <- mk_sc(YrefhatADT_norm, ADTtypeNames, refLabs)[,disNames]
sc_RNA <- mk_sc(PhiScRNA_norm, typeNames, quLabs)
sc_ADT <- mk_sc(PhiScADT_norm, ADTtypeNames, quLabs)

all(rownames(sc_RNA_ref) == rownames(sc_RNA))
rownames(sc_RNA_ref) <- rownames(sc_RNA) <- c(
  "ASDC", "B exhausted", "B immature", "B malignant",
  "B naive", "B unswitched", "B swithced", "C1 CD16 mono",
  "CD14 mono",       "CD16 mono",             "CD4 CM",                "CD4 EM",               
  "CD4 IL22",              "CD4 naive",             "CD4 prolif",            "CD4 Tfh",              
  "CD4 Th1",               "CD4 Th17",              "CD4 Th2",               "CD8 EM",               
  "CD8 naive",             "CD8 prolif",            "CD8 TE",                "CD83CD14 mono",       
  "DC prolif",             "DC1",                   "DC2",                   "DC3",                  
  "gdT",                   "HSC CD38-",           "HSC CD38+",           "HSC erythroid",        
  "HSC MK",                "HSC myeloid",           "HSC prolif",            "ILC1 3",               
  "ILC2",                  "MAIT",                  "mono prolif",           "NK CD16+",              
  "NK CD56+",               "NK prolif",             "NKT",                   "pDC",                  
  "plasma IgA",       "plasma IgG",       "plasma IgM",       "plasmablast",          
  "platelets",             "RBC",                   "Treg"                 
)
all(rownames(sc_ADT_ref) == rownames(sc_ADT))
rownames(sc_ADT_ref) <- rownames(sc_ADT) <- c(
  "B cell", "CD14 mono", "CD16 mono", "CD4 T", "CD8 T", "DC", "gdT",
  "HSC",          "lymph prolif",  "MAIT",         "mono prolif",  "NK CD16+",      "NK CD56+",      "pDC", 
  "plasmablast",  "platelets",    "RBC",          "Treg"
)

p1 <- tempCorHeat(
  cbind(
    sc_RNA_ref, sc_RNA
  )
)

p2 <- tempCorHeat(
  cbind(
    sc_ADT_ref, sc_ADT
  ),
  "ADT derived"
)

# pdf("corPlot_combo.pdf", width = 2.5, height = 6)
# draw(
#   p1 %v% p2,
#   heatmap_legend_side = "top"
# )
# dev.off()

p11 <- tempCorHeat(sc_RNA_ref, clust_rows = p1$clust_rows)$p
p21 <- tempCorHeat(
  sc_ADT_ref,
  "ADT derived",
  clust_rows = p2$clust_rows
)$plot
p12 <- tempCorHeat(sc_RNA, clust_rows = p1$clust_rows)$p
p22 <- tempCorHeat(
  sc_ADT,
  "ADT derived",
  clust_rows = p2$clust_rows
)$plot

pdf("figs/corPlot_combo.pdf", width = 2.5, height = 6)
draw(
  p11 %v% p21,
  heatmap_legend_side = "top"
)
dev.off()

pdf("figs/corPlot_combo_ADT_derived.pdf", width = 2, height = 6)
draw(
  p12 %v% p22,
  heatmap_legend_side = "top"
)
dev.off()


sc_combo <- cbind(
  rbind(sc_RNA_ref, sc_ADT_ref),
  rbind(sc_RNA, sc_ADT)
) %>% t()
clust_res <- hclust(dist(sc_combo))
pdf("figs/hclust.pdf", width = 2.5, height = 2.5)
ggdendro::ggdendrogram(clust_res)
dev.off()
```


Seurat V3 heatmaps: not meaningful since most scores are zero
```{r}
SeuRNAnew <- SeuRNA %>%
  `colnames<-`(
    gsub(
      "prediction.score.",
      "",
      colnames(SeuRNA)
    )
  )
SeuADTnew <- SeuADT %>%
  `colnames<-`(
    gsub(
      "prediction.score.",
      "",
      colnames(SeuADT)
    )
  )
SeuRNArefNew <- SeuRNAref %>%
  `colnames<-`(
    gsub(
      "prediction.score.",
      "",
      colnames(SeuRNAref)
    )
  )
SeuADTrefNew <- SeuADTref %>%
  `colnames<-`(
    gsub(
      "prediction.score.",
      "",
      colnames(SeuADTref)
    )
  )
sc_RNA_ref <- mk_sc(SeuRNArefNew, typeNames, refLabs)[,disNames]
sc_ADT_ref <- mk_sc(SeuADTrefNew, ADTtypeNames, refLabs)[,disNames]
sc_RNA <- mk_sc(SeuRNAnew, typeNames, quLabs)
sc_ADT <- mk_sc(SeuADTnew, ADTtypeNames, quLabs)


p1 <- tempCorHeat(
  cbind(
    sc_RNA_ref, sc_RNA
  )
)
p2 <- tempCorHeat(
  cbind(
    sc_ADT_ref, sc_ADT
  ),
  "ADT derived"
)


p11 <- tempCorHeat(sc_RNA_ref, clust_rows = p1$clust_rows)$p
p21 <- tempCorHeat(
  sc_ADT_ref,
  "ADT derived",
  clust_rows = p2$clust_rows
)$plot
p12 <- tempCorHeat(sc_RNA, clust_rows = p1$clust_rows)$p
p22 <- tempCorHeat(
  sc_ADT,
  "ADT derived",
  clust_rows = p2$clust_rows
)$plot

pdf("figs/corPlot_combo_Seu.pdf", width = 2.5, height = 6)
draw(
  p11 %v% p21,
  heatmap_legend_side = "top"
)
dev.off()

pdf("figs/corPlot_combo_ADT_derived_Seu.pdf", width = 2, height = 6)
draw(
  p12 %v% p22,
  heatmap_legend_side = "top"
)
dev.off()
```





# PLSDA (all but only show disease conds)

```{r}
XX <- cbind(PhiScADT_norm, PhiScRNA_norm)[quSubsetIdx,]
colnames(XX) <- c(
  paste0(colnames(PhiScADT_norm), "(ADT)"),
  paste0(colnames(PhiScRNA_norm), "(RNA)")
)
rownames(XX) <- rownames(PhiScADT_norm)[quSubsetIdx]
YY <- codeY(query, "Group")[rownames(XX), ]

ncomp <- 4
reg_res <- mvr(XX, YY, ncomp = ncomp, method = "PLS", center = T, scale = F)

Bhat <- reg_res$coefficients[,,ncomp]
```

Plot

```{r fig.asp=1}
p <- plotBmat(Bhat)
ggsave(
  "figs/feat2autoimmune.pdf", adjPlots2(p, "bw"), width = 2.5, height = 4
)
```










