---
title: "Case DC: projecting scRNA-seq to bulk RNA-seq atlas"
output: html_notebook
---



Note that `dat_dir` is the directory where you store the downloaded data and output from the R scripts. This directory should have the following structure:
```
dat_dir
│   
└───data
│   │
│   └───stemformatics
│   |   │   ref_dc.rds
│   |   │   ...
|   |   
│   └───Covid
|   |...
|   |
└───output
    │  
    └───CaseDC
    |
    └───CaseCITE
    |
    └───CaseATAC
```

```{r}
suppressPackageStartupMessages(library(PhiSpace))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(ComplexHeatmap))

dat_dir <- "/home/unimelb.edu.au/jmao1/PhiSpace/" # Replace this by your own directory 
source("Rosa_cols.R") # colour code for visualising 
```

Load data. The processed reference and query data can be downloaded from [OneDrive](https://unimelbcloud-my.sharepoint.com/:f:/g/personal/jiadong_mao_unimelb_edu_au/EimPelXY7CpHqNKd4vKfLQABhDNcmkpQtzlm2uKOQMGtPQ?e=uQhpBR). See `processData.R` for creating these data objects.

**Warning** This session may require up to 50GB RAM. If you don't have access to bigger RAM, consider subsetting the query datasets (eg to 5000 cells).
```{r}
query <- readRDS(paste0(dat_dir, "data/Rosa/query_Rosa.rds"))
# Comment the following line if you do not wish to subset
query <- subsample(query, key = "mainTypes", proportion = 0.1, minCellNum = 50)
reference <- readRDS(paste0(dat_dir,"data/stemformatics/ref_dc.rds"))
```

In this case study we use rank normalisation for both reference and query. This is because the reference was rank normalised; see [this paper](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008219) for an explanation. 
```{r}
query <- RankTransf(query, "counts")
reference <- RankTransf(reference, "data", sparse = F)
```


Annotate query cells. See `CaseDC_tune.R` for feature selection.
```{r}
PhiSpaceAssay <- "rank"
phenotypes <- c("Cell Type", "Sample Source")
PhiMethod <- "PLS"

PhiResPath <- paste0(
  dat_dir,
  "output/CaseDC/PhiRes.rds"
)

if(!file.exists(PhiResPath)){
  
  selectedFeat <- readRDS(paste0(dat_dir, "output/CaseDC/ref_dc_test.rds"))
  PhiRes <- PhiSpaceR_1ref(
    reference, 
    query,
    ncomp = NULL,
    selectedFeat = selectedFeat,
    phenotypes = phenotypes, 
    PhiSpaceAssay = PhiSpaceAssay,
    regMethod = PhiMethod,
    scale = FALSE, 
    DRinfo = F
  )
  
  saveRDS(PhiRes, PhiResPath)
} else {
  
  PhiRes <- readRDS(PhiResPath)
}
```
 

Use normalised PhiSpace scores for downstream analyses.
```{r}
YrefHat <- PhiRes$YrefHat
YrefHat_norm <- normPhiScores(YrefHat)
idx <- colnames(query)
PhiScores_norm <- normPhiScores(PhiRes$PhiSpaceScore)[idx,]
```


Heatmap.
```{r}
queryLabs <- query$mainTypes
queryLabs[queryLabs %in% c("Day9_SP", "Day9_DP")] <- "Day9"
lvls <- c("DC1", "DC2", "pDC", "HEF", "Day3", "Day6", "Day9")
p <- plotPhiSpaceHeatMap(
  PhiSpaceScore = PhiScores_norm,
  reference = reference,
  phenotypes = phenotypes,
  queryLabs = queryLabs, 
  queryLvls = lvls,
  column_names_rot = 20,
  name = "Phenotype space embedding",
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 6),
  show_row_dend = F,
  show_column_dend = T,
  # row_title = row_title, 
  row_title_gp = gpar(fontsize = 6),
  column_title_gp = gpar(fontsize = 6, fontface = "bold"),
  heatmap_legend_param = list(
    title_position = "leftcenter",
    title_gp = gpar(fontsize = 6),
    grid_height = unit(2, "mm"),
    grid_width = unit(2, "mm"),
    labels_gp = gpar(fontsize = 5),
    legend_direction = "horizontal"
  )
) 

pdf("figs/heatmap.pdf", width = 6, height = 5)
draw(
  p,
  heatmap_legend_side = "top"
)
dev.off()
```


Simplify labels for better visualisation.
```{r}
queryLabs <- query$mainTypes
queryLabs[queryLabs %in% c("Day9_SP", "Day9_DP")] <- "Day9"
refLabs <- colData(reference)[,"Cell Type"]
allSymbs <- c(
  colData(reference)[,"Sample Source"],
  rep("query", nrow(PhiScores_norm))
)
allLabs <- c(refLabs, queryLabs)
# allLabs[allLabs %in% c("pDC", "DC2")] <- "other"
allLabs <- factor(
  allLabs,
  levels = c("DC", "MoDC", "DC_prec", "mono", "pDC",
             "DC1", "DC2", "HEF",
             "Day3", "Day6", "Day9")
)
```




Phenotype PC space.

```{r}
pc_re <- getPC(YrefHat_norm, ncomp = 3)
plot_dat <- rbind(
  pc_re$scores,
  scale(
    PhiScores_norm, 
    center = T, 
    scale = F
  ) %*% pc_re$loadings
) %>%
  as.matrix() %>%
  as.data.frame() %>%
  mutate(
    `Cell type` = allLabs,
    `Sample source` = allSymbs,
    `Dataset` = rep(c("reference", "query"), c(nrow(pc_re$scores), nrow(PhiScores_norm)))
  )
```

A function for adjusting themes of your ggplot2 object for more compact presentation, useful for papers.
```{r}
ref_size <- 1
query_size <- 1.5
query_stroke <- 0.05
axis.title.x.margin <- -2
axis.title.y.margin <- -3

adjPlots <- function(
  p,
  theme_overall = c("bw", "ggplot2", "classic", "linedraw", "light", "dark", "minimal", "void"),
  # font sizes
  default.fsize = 10,
  axis.text.x.fsize = NULL,
  axis.text.y.fsize = NULL,
  axis.title.x.fsize = NULL,
  axis.title.y.fsize = NULL,
  legend.text.fsize = NULL,
  legend.title.fsize = NULL,
  # margins
  axis.title.x.margin = 0,
  axis.title.y.margin = 0,
  # positions
  legend.position = "top",
  legend.title.position = "left",
  # spacings
  legend.spacing.x = -5,
  legend.key.spacing = -5,
  legend.key.spacing.y = -10,
  legend.box.margin = 0,
  legend.box.spacing = 0,
  # sizes
  legend.obj.size = 2
){
  
  
  theme_overall <- match.arg(theme_overall)
  theme_fun <- switch (
    theme_overall,
    classic = theme_classic,
    ggplot2 = theme_gray,
    bw = theme_bw,
    linedraw = theme_linedraw,
    light = theme_light,
    dark = theme_dark,
    minimal = theme_minimal,
    void = theme_void
  )
  
  ## Font sizes
  if(is.null(axis.text.x.fsize)) axis.text.x.fsize <- default.fsize
  if(is.null(axis.text.y.fsize)) axis.text.y.fsize <- default.fsize
  if(is.null(axis.title.x.fsize)) axis.title.x.fsize <- default.fsize
  if(is.null(axis.title.y.fsize)) axis.title.y.fsize <- default.fsize
  if(is.null(legend.text.fsize)) legend.text.fsize <- default.fsize
  if(is.null(legend.title.fsize)) legend.title.fsize <- default.fsize
  
  p_adj <- p +
    theme_fun() +
    theme(
      axis.text = element_text(size = axis.text.x.fsize),
      axis.title.x = element_text(
        size = axis.title.x.fsize, 
        margin = margin(
          axis.title.x.margin, 
          axis.title.x.margin, 
          axis.title.x.margin, 
          axis.title.x.margin
        )
      ),
      axis.title.y = element_text(
        size = axis.title.y.fsize, 
        margin = margin(
          axis.title.y.margin,
          axis.title.y.margin,
          axis.title.y.margin,
          axis.title.y.margin
        )
      ),
      legend.text = element_text(size = legend.text.fsize),
      legend.title = element_text(size = legend.title.fsize, face = "bold"),
      legend.position = legend.position,
      legend.title.position = legend.title.position,
      legend.key = element_blank(),
      legend.spacing.x = unit(legend.spacing.x, "pt"),
      legend.key.spacing = unit(legend.key.spacing, "pt"),
      legend.key.spacing.y = unit(legend.key.spacing.y, "pt"),
      legend.box.margin = margin(
        legend.box.margin,
        legend.box.margin,
        legend.box.margin,
        legend.box.margin
      ),
      legend.box.spacing = unit(legend.box.spacing, "pt"),
      legend.background = element_blank()
    ) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = legend.obj.size
        ),
        nrow = 1
      ),
      shape = guide_legend(
        override.aes = list(
          size = legend.obj.size
        )
      )
    )
  
}

tempPlot <- function(coord1 = "comp1", coord2 = "comp2"){
  
  p <- 
    plot_dat %>%
    ggplot(
      aes(x = plot_dat[,coord1], y = plot_dat[,coord2])
    ) +
    geom_point(
      aes(colour = `Cell type`, shape = Dataset, size = Dataset),
      stroke = query_stroke
    ) +
    scale_size_manual(values = c(query_size, ref_size)) +
    scale_colour_manual(
      values = DC_cols,
      breaks = c("DC1", "DC2", "HEF",
                 "Day3", "Day6", "Day9", 
                 "DC", "DC_prec", "mono", 
                 "pDC", "DC2", "MoDC", "other")
    ) +
    scale_shape_manual(values = DC_symbs) +
    labs(x = coord1, y = coord2)
  
  p <- adjPlots(p)
  
  return(p)
}

tempPlot_source <- function(coord1 = "comp1", coord2 = "comp2"){
  
  p <- 
    plot_dat %>%
    ggplot(
      aes(x = plot_dat[,coord1], y = plot_dat[,coord2])
    ) +
    geom_point(
      aes(colour = `Sample source`, shape = Dataset, size = Dataset),
      stroke = 0.1
    ) +
    theme_bw() +
    scale_size_manual(
      values = c(query_size, ref_size)
    ) +
    scale_colour_manual(
      values = DC_cols_source
    ) +
    scale_shape_manual(values = DC_symbs) +
    labs(x = coord1, y = coord2) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = 2,
          shape = c(
            rep(16, 4),
            4
          )
        ),
        nrow = 2
      ),
      size = "none",
      shape = "none"
    ) +
    theme(
      axis.text = element_text(size = 6),
      axis.title.x = element_text(
        size = 6, 
        margin = margin(
          axis.title.x.margin, 
          axis.title.x.margin, 
          axis.title.x.margin, 
          axis.title.x.margin
        )
      ),
      axis.title.y = element_text(
        size = 6, 
        margin = margin(
          axis.title.y.margin,
          axis.title.y.margin,
          axis.title.y.margin,
          axis.title.y.margin
        )
      ),
      legend.text = element_text(size = 6),
      legend.title = element_text(size = 6, face = "bold"),
      legend.position = "top",
      legend.key = element_blank(),
      legend.spacing.x = unit(-5, "pt"),
      legend.key.spacing = unit(-5, "pt"),
      legend.key.spacing.y = unit(-10, "pt"),
      legend.title.position = "top",
      legend.box.margin = margin(-10, -10 ,-10 , -10),
      legend.box.spacing = unit(0, "inch"),
      legend.background = element_blank()
    )
  
  # theme_void 
  # angle of x and y labels
  
    
  
  return(p)
}

# Comp 1 & 2
p_comp12 <- tempPlot()
p_comp23 <- tempPlot("comp2", "comp3")
p_comp13 <- tempPlot("comp1", "comp3")

p_comp12_source <- tempPlot_source()
p_comp23_source <- tempPlot_source("comp2", "comp3")
p_comp13_source <- tempPlot_source("comp1", "comp3")

ggsave(
  filename = "figs/Query2PhenoSpacePC.pdf",
  plot = p_comp12, 
  width = 7.5,
  height = 2.5
)
ggsave(
  filename = "figs/Query2PhenoSpacePC_sampSource.pdf",
  plot = p_comp12_source, 
  width = 2.5,
  height = 2.5
)
ggsave(
  filename = "figs/Query2PhenoSpacePC23.pdf",
  plot = p_comp23, 
  width = 2.5,
  height = 2.5
)
ggsave(
  filename = "figs/Query2PhenoSpacePC23_sampSource.pdf",
  plot = p_comp23_source, 
  width = 2.5,
  height = 2.5
)
ggsave(
  filename = "figs/Query2PhenoSpacePC13.pdf",
  plot = p_comp13, 
  width = 2.5,
  height = 2.5
)
ggsave(
  filename = "figs/Query2PhenoSpacePC13_sampSource.pdf",
  plot = p_comp13_source, 
  width = 2.5,
  height = 2.5
)
```

Biplot overlaid.

```{r}
tempPlot_load <- function(comps = c(1,2)){
  
  comps1 <- paste0("comp", comps[1])
  comps2 <- paste0("comp", comps[2])
  
  plot_load <- 
    pc_re$loadings %>%
    as.matrix() %>%
    as.data.frame() %>%
    mutate(
      importance = (!!sym(comps1))^2 + (!!sym(comps2))^2,
      names = rownames(.)
    ) %>%
    arrange(-importance)
  
  p <- plot_dat %>%
    filter(Dataset == "query") %>%
    ggplot(aes(x = !!sym(comps1), y = !!sym(comps2))) +
    geom_point(aes(colour = queryLabs), shape = 4, size = 1.5, stroke = 0.3) +
    theme_bw() +
    scale_colour_manual(
      values = DC_cols,
      breaks = c(
        "HEF", "Day3", "Day6", "Day9",
        "DC1", "DC2", "pDC"
      )
    ) +
    geom_segment(
      data = plot_load,
      aes(x = 0, y = 0, xend = (!!sym(comps1)) * 0.9, yend = (!!sym(comps2)) * 0.9),
      arrow = arrow(angle = 10, length = unit(3, "points")),
      linewidth = 0.3
    ) +
    geom_label(
      data = plot_load,
      aes(
        label = names,
        size = names
      )
    ) +
    scale_size_manual(
      values = c(
        DC1 = 2,
        DC = 1,
        in_vitro = 2,
        ex_vivo = 2,
        DC2 = 2,
        in_vivo_HuMouse = 1,
        MoDC = 1,
        mono = 1,
        in_vivo = 2,
        pDC = 1,
        DC_prec = 1
      ),
      guide = "none"
    ) +
    theme(
      axis.text = element_text(size = 6),
      axis.title.x = element_text(
          size = 6, 
          margin = margin(
            axis.title.x.margin, 
            axis.title.x.margin, 
            axis.title.x.margin, 
            axis.title.x.margin
          )
        ),
        axis.title.y = element_text(
          size = 6, 
          margin = margin(
            axis.title.y.margin,
            axis.title.y.margin,
            axis.title.y.margin,
            axis.title.y.margin
          )
        ),
      legend.text = element_text(size = 6),
      legend.title = element_text(size = 6),
      legend.position = "top",
      legend.key = element_blank(),
      legend.spacing.x = unit(-5, "pt"),
      legend.key.spacing = unit(-5, "pt"),
      legend.key.spacing.y = unit(-10, "pt"),
      legend.title.position = "top", 
      legend.box.margin = margin(-10,-10,-10,-10),
      legend.box.spacing = unit(0, "inch"),
      legend.background = element_blank()
    ) +
    guides(
      colour = guide_legend(
        title = "Cell type"
      )
    )
  
  return(p)
}

p12 <- tempPlot_load()
p23 <- tempPlot_load(c(2,3))

ggsave(
  filename = "figs/biplot.pdf",
  plot = p12,
  width = 2.5,
  height = 2.5
)
ggsave(
  filename = "figs/biplot23.pdf",
  plot = p23,
  width = 2.5,
  height = 2.5
)
```




# Compare with SingleR

See `Case1_Rosa_SingleR.R` for getting SingleR annotation.

```{r}
SingleRScores <- readRDS(paste0(dat_dir, "output/Case1/Rosa2dc_SR_score.rds"))
SingleRScores_noselect <- readRDS(paste0(dat_dir, "output/Case1/Rosa2dc_SR_score_noselect.rds"))
SingleRScores_stemf <- readRDS(paste0(dat_dir, "output/Case1/Rosa2dc_SR_score_stemf_feat.rds"))

queryLabs <- query$mainTypes
queryLabs[queryLabs %in% c("Day9_SP", "Day9_DP")] <- "Day9"
lvls <- c("DC1", "DC2", "pDC", "HEF", "Day3", "Day6", "Day9")

PhiScores_norm_mod <- PhiScores_norm
newColNames <- colnames(PhiScores_norm)
newColNames[3] <- "cDC1"
colnames(PhiScores_norm_mod) <- newColNames
```

Violin plots.
```{r}
temp_plot_violin <- function(scores, title){
  p <-
    scores %>%
    as.data.frame() %>%
    mutate(queryLabs = factor(queryLabs, levels = lvls)) %>%
    pivot_longer(
      cols = ! queryLabs, 
      names_to = "refLabs", 
      values_to = "scores"
    ) %>%
    filter(refLabs %in% c("cDC1", "in_vitro")) %>%
    ggplot(aes(x = queryLabs, y = scores, fill = refLabs)) +
    geom_violin() +
    theme_bw() +
    scale_fill_discrete(
      labels = c(
        "DC1", "in_vitro"
      )
    ) +
    guides(fill = guide_legend(title = "Phenotype")) +
    ggtitle(title) +
    theme(
      axis.text = element_text(size = 6),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 7, face = "bold"),
      legend.text = element_text(size = 6),
      legend.title = element_text(size = 6),
      legend.position = "top",
      # legend.key = element_blank(),
      legend.background = element_blank(),
      legend.key.size = unit(6, "pt")
    ) 
    
  return(p)
}
## Boxplots for PhiSpace scores
p_phi <- temp_plot_violin(PhiScores_norm_mod, "PhiSpace")
p_sr <- temp_plot_violin(SingleRScores, "SingleR")
p_sr_ns <- temp_plot_violin(SingleRScores_noselect, "SingleR using all features")
p_sr_stemf <- temp_plot_violin(SingleRScores_stemf, "SingleR using features selected by Elahi et al. (2022)")
# ggarrange(p_phi, p_sr_ns, p_sr, ncol = 1, common.legend = TRUE)
ggsave(
  "figs/violin.pdf",
  plot = ggarrange(
    p_phi, 
    # p_sr_ns, 
    p_sr, 
    # p_sr_stemf,  
    nrow = 1, ncol = 2, common.legend = TRUE
  ),
  width = 7,
  height = 2
)
```



# Omics space reference mapping
```{r}
selectedFeat <- rownames(PhiRes$atlas_re$reg_re$loadings)

assay4pc <- t(
    assay(reference, PhiSpaceAssay)[selectedFeat,]
  )
assay4pc <- PhiSpace:::RTassay(assay4pc)

pca_omics_res <- getPC(
  assay4pc,
  ncomp = 10
)

plot_dat <- rbind(
  pca_omics_res$scores,
  scale(
    t(assay(query, PhiSpaceAssay))[,selectedFeat], 
    center = T, 
    scale = F
  ) %*% pca_omics_res$loadings
) %>%
  as.matrix() %>%
  as.data.frame() %>%
  mutate(
    `Cell type` = allLabs,
    `Sample source` = allSymbs,
    `Dataset` = rep(
      c("reference", "query"), 
      c(
        nrow(pca_omics_res$scores), 
        nrow(PhiScores_norm)
      )
    )
  )
```



```{r}
# Comp 1 & 2
p_comp12 <- tempPlot()
p_comp23 <- tempPlot("comp2", "comp3")
p_comp13 <- tempPlot("comp1", "comp3")

p_comp12_source <- tempPlot_source()
p_comp23_source <- tempPlot_source("comp2", "comp3")
p_comp13_source <- tempPlot_source("comp1", "comp3")
  

ggsave(
  "figs/Xspace.pdf",
  plot = p_comp12,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/Xspace_source.pdf",
  plot = p_comp12_source,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/XspacePC23.pdf",
  plot = p_comp23,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/Xspace_sourcePC23.pdf",
  plot = p_comp23_source,
  width = 2.5,
  height = 2.5
)
```


Sincast imputation. See `Case_DC_runSincast.R` for details. Took 25.50501 mins to compute with 4 CPUs and 250GB RAM.
```{r}
SinResPath <- paste0(
  dat_dir,
  "output/Case1/SinImputRosa.rds"
)

querySin <- readRDS(SinResPath)
querySin <- RankTransf(querySin, "SincastImpData")
```

```{r}
plot_dat <- rbind(
  pca_omics_res$scores,
  scale(
    t(assay(querySin, PhiSpaceAssay))[,selectedFeat], 
    center = T, 
    scale = F
  ) %*% pca_omics_res$loadings
) %>%
  as.matrix() %>%
  as.data.frame() %>%
  mutate(
    `Cell type` = allLabs,
    `Sample source` = allSymbs,
    `Dataset` = rep(
      c("reference", "query"), 
      c(
        nrow(pca_omics_res$scores), 
        nrow(PhiScores_norm)
      )
    )
  )
```

```{r}
# Comp 1 & 2
p_comp12 <- tempPlot()
p_comp23 <- tempPlot("comp2", "comp3")
p_comp13 <- tempPlot("comp1", "comp3")

p_comp12_source <- tempPlot_source()
p_comp23_source <- tempPlot_source("comp2", "comp3")
p_comp13_source <- tempPlot_source("comp1", "comp3")
  

ggsave(
  "figs/XspaceSincast.pdf",
  plot = p_comp12,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/XspaceSincast_source.pdf",
  plot = p_comp12_source,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/XspaceSincastPC23.pdf",
  plot = p_comp23,
  width = 2.5,
  height = 2.5
)
ggsave(
  "figs/XspaceSincast_sourcePC23.pdf",
  plot = p_comp23_source,
  width = 2.5,
  height = 2.5
)
```


PC plots.
```{r}
## Colour scheme
cols <- c(
  DC_prec = "#B3B3B3",
  MoDC = "#CCEBC5", 
  DC = "#B3DE69", 
  mono = "#80B1D3",
  DC1 = "#1B9E77",
  DC2 = "#D95F02",
  pDC = "#7570B3",
  # Sample source
  ex_vivo = "#B3B3B3",
  in_vitro = "#A50026",
  in_vivo = "#313695",
  `in_vivo_HuMouse` = "#7570B3",
  # platform
  HuGene = "#8DD3C7",
  Illumina = "#FFFFB3",
  `Illumina V4` = "#BEBADA",
  RNAseq = "#FB8072"
)

# Batch and biology
phes <- c("Platform Category", "Cell Type", "Sample Source")
legendTitle <- c("Platform", "Cell type", "Sample source")
outPlots <- vector("list", length(phes))
for(i in 1:length(phes)){
  phe <- phes[i]
  
  outPlots_i <- vector("list", length(pc_list))
  for(j in 1:length(pc_list)){
    
    outPlots_i[[j]] <- pc_list[[j]]$scores %>%
      as.matrix() %>%
      as.data.frame() %>%
      mutate(labs = colData(reference)[,phe]) %>%
      ggplot(aes(x = comp1, y = comp2)) +
      geom_point(
        aes(colour = labs),
        size = 0.5
      ) + 
      theme_bw() +
      scale_colour_manual(
        values = cols
      ) +
      theme(
        axis.title = element_blank(),
        axis.text = element_text(size = 6),
        # plot.title = element_text(size = 7, face = "bold"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.position = "bottom",
        legend.key = element_blank(),
        legend.background = element_blank(),
        legend.key.size = unit(8, "pt"),
        legend.box.margin = margin(-5,-5,-5,-5),
        legend.box.spacing = unit(0, "inch")
      )  +
      guides(
        colour = guide_legend(
          title = legendTitle[i],
          nrow = 1,
          override.aes = list(
            size = 2
          )
        )
      )
  }
  
  outPlots[[i]] <- ggarrange(
    plotlist = outPlots_i, 
    nrow = 1, 
    common.legend = TRUE, 
    legend = "bottom"
  )
}

ggsave(
  "figs/CompareFeatSelect.pdf",
  plot =  ggarrange(plotlist = outPlots, ncol = 1),
  width = 7,
  height = 6.5
)
```